\documentclass[parskip=full]{scrartcl}

\pdfoutput=1

\title{Imbalanced Learning in Land Cover Classification  \\ \LARGE{Improving minority class' prediction accuracy using the Geometric SMOTE algorithm}}

\author{
	Georgios Douzas\(^{1}\), Fernando Bacao\(^{1*}\), Joao Fonseca\(^{1}\), Manvel Khudinyan\(^{1}\)
	\\
	\small{\(^{1}\)NOVA Information Management School, Universidade Nova de Lisboa}
	\\
	\small{*Corresponding Author}
	\\
	\\
	\small{Postal Address: NOVA Information Management School, Campus de Campolide, 1070-312 Lisboa, Portugal}
	\\
	\small{Telephone: +351 21 382 8610}
}

\usepackage{breakcites}
\usepackage{float}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{geometry}
\geometry{
	a4paper,
	total={170mm,257mm},
	left=18mm,
	right=18mm,
	top=8mm,
}
\usepackage{amsmath}
\newcommand{\inlineeqnum}{\refstepcounter{equation}~~\mbox{(\theequation)}}
\usepackage{enumitem}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{booktabs}
\usepackage{pgfplotstable}
\pgfplotsset{compat=1.14}
\usepackage{longtable}
\usepackage{tabu}
\usepackage{hyperref}
\usepackage{csvsimple}
\date{}

\begin{document}

\maketitle

\begin{abstract}
Abstract goes here
\end{abstract}

\section{Introduction}

The production of accurate Land Use/Land Cover maps offer unique monitoring
capabilities within the remote-sensing domain \cite{Mellor2015}. LULC maps are
being used for a variety of applications, ranging from environmental
monitoring, land change detection, natural hazard assessment up to agriculture
and water/wetland monitoring \cite{Khatami2016}.

Multispectral images are an important resource to build LULC maps, allowing for
the use of classification algorithms, both supervised and unsupervised, to
automate the production of these maps. Although significant progress has been
made in the use of supervised learning techniques for automatic image
classification \cite{Tewkesbury2015}, the acquisition of labeled training sets
continues to be a bottleneck \cite{Rajan2008}. To build  good and robust
supervised classifiers it is crucial to have a large enough training dataset.
Often, the problem is that different land use classes have very different
levels of coverage, while some are frequent in the training dataset, other can
be limited \cite{Feng2019}. In remote-sensing classification problems, the
number of instances of some classes is often very small when compared with the
dominant classes, this is especially true in LULC data \cite{Williams2009,
Cenggoro2018}.

This imbalance in the representation of the different classes constitutes a
problem for the development of accurate classifiers. The problem is usually
referred to, in the machine learning community, as the imbalanced learning
problem \cite{Chawla2004}. The imbalanced learning problem is even more
challenging in multi-class problems, making it one of the most
important challenges in the machine learning research community
\cite{Garcia2018}. Improvements in the representation of small classes in
training datasets can have a significant impact in improving the accuracy and
robustness of classifiers for the production of LULC maps.

An imbalanced learning problem refers to a skewed distribution of data across
classes in both binary and multi-class classification problems \cite{Abdi2016}.
In this situation, the minority classes contribute less to the minimization of
accuracy, the typical objective function, inducing a bias towards the majority
class during the learning process \cite{Douzas2019}. Consequently, as typical
classifier learning methods are designed to work with reasonably balanced
datasets, finding meaningful boundaries between the different classes becomes a 
very difficult task\cite{Saez2016}.

Possible approaches to deal with class imbalance can be divided into three main
groups \cite{Fernandez2013}. 1) Cost-sensitive solutions: adaptations at
algorithmic and/or data level by applying higher misclassification costs for the
examples of the minority class. 2) Algorithmic level solutions: classification
algorithms are adapted or created to reinforce the learning of the positive
class. 3) Data level solutions: a more general approach where class
distributions are rebalanced through the sampling of the data space, thus
diminishing class imbalance. Because this latter method is more general, it is
used for a wide range of applications, thus making it of particular interest.
Regardless, data level solutions carry the disadvantage of yielding an increased
search space.

There are several data level solutions to deal with the imbalanced learning
problem, which can be divided into three groups. While Undersampling algorithms
reduce the size of the majority classes, oversampling algorithms attempt to
even the distributions by generating artificial data for the minority class
\cite{Mellor2015}. A hybrid approach, on the other hand, uses both oversampling
and undersampling techniques to ensure a balanced dataset.

Studies suggest that the usage of data level solutions in remote sensing
problems, to balance class distribution, seem to outperform models
trained by randomly drawn samples \cite{Wang2019, Mellor2015}.  Studies
employing oversampling methods such as the Synthetic Minority Over-sampling
Technique (SMOTE) \cite{Chawla2002} seem to consistently yield better results
\cite{Johnson2013, Geib2015} when compared to no oversampling. Other studies 
employ active learning methods such as Margin Sample by Closest Support Vector
(MS-cSV), as it benefits from avoiding oversampling in dense regions close to
the margin and samples all the feature space equivalently \cite{Tuia2009}.
In spite of the potential benefits that oversampling techniques can bring to 
the improvement of automatic classification, there are not many studies on the 
subject in the remote sensing domain.

In this paper, we test the performance of oversampling techniques on an
imbalanced dataset using different types of classifiers. The effectiveness of
the Geometric SMOTE (G-SMOTE) \cite{Douzas2019} is demonstrated using SMOTE
\cite{Chawla2002}, Borderline SMOTE \cite{Han2005}, ADASYN \cite{HaiboHe2008},
Random Oversampling and no oversampling as baseline methods. Preliminary results
show that G-SMOTE outperforms every other oversampling technique, for both
overall prediction power and minority class prediction power.

Oversamplers' performance is evaluated through an experimental analysis using
the publicly available LUCAS dataset \cite{LUCAS2015}. This dataset contains
multispectral reflectance data from Landsat 8, filtered for a region in
Portugal. The experimental procedure includes a comparison of the different
oversamplers using 4 classifiers and \textbf{(...)} evaluation metrics.

This paper is organized in \textbf{(...)} sections: \textbf{(...)}

\section{State of the Art}

Data modification through resampling has been the most popular approach to deal 
with the imbalanced learning problem, in machine learning in general 
\cite{Douzas2019} and remote sensing in particular \cite{Feng2019}. By 
decoupling the imbalance problem from the classification algorithms, resampling 
allows the users to apply any standard algorithm once the resampling 
preprocessing step is done. This is especially convenient for users that are 
not machine learning experts and want to use several classifiers. Additionally, 
resampling methods can be easily adapted to multi-class imbalanced data, which 
is relevant for LULC classification \cite{Feng2019}. In this section we present 
the most relevant applications of resampling methods for remote sensing 
imbalanced data classification. 

Some of the existing studies implements the random undersampling method,
which randomly reduces the number of training samples  belonging to the
majority class. However, this method has the disadvantage of information loss
as it discards samples from the majority class  \cite{Feng2019}.
\cite{Feng2018} carries out competitive analysis of  undersamplers and
oversamplers performance for land cover classification with  Rotation Forest
(RoF) ensemble classifier. In the experiment random  undersampling,
undersampling based RoF and number of oversampling techniques  were deployed.
The dataset used comprises 16 imbalanced classes from  hypercritical imagery.
The results reported show the limited interest of the  undersampling methods as 
they always produce worst results in the prediction of all the 16 classes.

Contrary to random undersampling, the random oversampling method avoids 
information loss. However, this method simply replicates randomly selected 
instances of the minority class, consequently, increasing the risk of 
over-fitting \cite{Krawczyk2016}. \cite{Maxwell2018} reports that balancing 
data with random oversampling affects the classification performance 
differently for different supervised models. In their paper land cover 
classification with highly imbalanced data is carried out with six different 
supervised classifiers. Implementation of random oversampling could slightly 
improve the performance of Random Forest (RF) and Support Vector Machine (SVM) 
classifiers. On the other hand it reduced the overall classification accuracy 
for classifiers such as Decision Tree (DT), Artificial Neural Network (ANN), 
k-Nearest Neighbors (k-NN) and Boosted DT. The paper shows, that even if random 
oversampling can positively impact on the prediction of minority classes, 
also it can reduce the accuracy in the majority class.

Synthetic minority oversampling technique (SMOTE) \cite{Chawla2002} is the most 
popular oversampling method, and it has been used to successfully deal with the 
class imbalance problem in land cover classification. The variational 
semi-supervised learning (VSSL) proposed by \cite{Cenggoro2018} aims to fix the 
imbalance problem in LULC mapping. VSSL is a semi-supervised learning framework 
consisting of a deep generative model allowing learning from both labeled 
and unlabeled samples while using SMOTE to balance the data. The results show 
significant improvements in performance on the producer’s accuracy of the 
minority class, when compared to the results without any imbalanced learning. 

Ensemble learning method together with oversampling algorithms have been
successfully implemented for remote sensing image classification in the resent
years. \cite{Feng2018}, \cite{Feng2019} apply the Rotation Forest ensemble 
learning algorithm with SMOTE to deal with class imbalance problem. Those 
papers offer a novel approach to the use of SMOTE based RoF algorithm for a 
hyperspectral image classification. The idea is to iteratively balance the 
class distribution using SMOTE before creating each rotation decision tree. 
The experiments results yields about improved classification accuracy in the 
minority classes,  as well as for the whole dataset.

A number of studies report significant improvements in LULC mapping accuracy
with the use of SMOTE oversampling along with standard algorithms to handle 
with data imbalance problem for multi-source remote sensing data. 
\cite{Johnson2016} uses OpenStreetMap crowdsourced data and Landsat time series 
for LULC classification. A simple implementation of SMOTE oversampler shows 
improved classification results for all the supervised classifiers (naive 
bayes, DT and RF). \cite{Bogner2018} use RF classifier to detect rare land use 
classes on MODIS time series. They use SMOTE to decrease the imbalance ratio
in the original dataset. The comparison of different classification scenarios 
shows the minority classes to be better classified after SMOTE. 
\cite{Panda2018} carried out a simple implementation of SMOTE with five 
different classifiers (Nave Bayes, DT, RF, k-NN, SVM) to predict a land cover 
maps with 4, 5 and 6 classes. The results show the relevance of using 
oversampling to improve the classifiers performance, as 5 of the classifiers 
improve their results.

Even if recent studies demonstrate the usefulness of oversampling methods
for remote sensing applications, they still present some drawbacks. Random
oversampling often leads to the model over-fitting as it simply replicates the
instances for minority class \cite{Feng2019}. SMOTE method can avoid
over-fitting but has the disadvantage of generating noisy data \cite{He2008}.
In order to mitigate this problem many variations of SMOTE has been developed
with the objective of preventing noisy data generation (see section 1). In this 
study we use a set of more recent oversamplers, most notably Geometric-SMOTE, 
that are yet to be adopted by remote sensing community. 

A novel oversampling technique Geometric-SMOTE algorithm is an extension of 
SMOTE algorithm and has been proposed recently by \cite{Douzas2019}. Similar to 
SMOTE, the size of the area for data generation for Geometric-SMOTE is derived 
form the distance of the selected minority sample to the nearest neighbor using 
$k$-neighbors. The main difference to SMOTE algorithm is that instead of using 
line segment between minority samples, Geometric-SMOTE defines a flexible 
geometric region (hyper-spheroid) around each minority class instance for 
synthetic data generation. The shape of geometric regions are defined with 
$deformation factor$ and $truncation factor$ hyper-parameters. This intensively 
increases the options for more informative data generation. Further, 
Geometric-SMOTE is designed to avoid noisy sample generation by some 
restrictions in the neighbor selection strategy.

Borderline-SMOTE is one of the most popular SMOTE based ovrsamplers. It has two 
variations, Borderline-SMOTE1 and Borderline-SMOTE2 and focuses on the 
oversampling only the minority instances near the borderline to create better 
boundaries between classes \cite{Han2005}. As in the case of SMOTE, it 
calculates $k$ nearest neighbors for all the minority instances. Next, only 
borderline minority samples are selected as DANGER set if the majority of the 
$k$ nearest neighbors belong to the majority class. Borderline-SMOTE1 method 
uses all instances from DANGER set to calculates $s$ nearest neighbors from 
minority class and generate synthetic instances with a simple SMOTE method. 
Borderline-SMOTE2 uses not only $s$ minority neighbors for each instance of 
DANGER set, but also the nearest majority class neighbors. For this, the 
difference between the DANGER set instance and majority class instance is 
multiplied with a random number between 0 and 0.5 to ensure that the new 
generated instance is located closer to the minority class. 

The Adaptive Synthetic Sampling Technique (ADASYN) is another well known 
oversampling method based on SMOTE. Adaptive Synthetic sampling is based on the 
idea of adaptively generating minority class instances according to their 
weighted distribution: more instances are generated for those minority class 
instances that are harder to learn compared to ones that are easier to learn 
\cite{HaiboHe2008}. With this characteristics ADASYN differs from SMOTE method, 
which gives equal chances for all the minority instances to be selected for 
synthetic data generation. The synthetic instances are created based on the 
majority nearest neighbors using $k$ nearest neighbor method. ADASYN 
oversampling method has the advantage of reducing the learning bias introduced 
by the original imbalance data distribution. One drawback of this approach is 
that it does not identify noisy instances, and thus it does not prevent from 
adding outlier values in the dataset.

To compare the performance of the different oversamplers we use the LUCAS 
dataset and 4 different supervised classifiers: k-nearest neighbours, decision 
trees, random forest and gradient boosting classifier. \todo{exists also in the 
introduction}

\section{Methodology}

This section explains the evaluation process of G-SMOTE's performance. A
description of the baseline methods, classification algorithms, evaluation
methods, dataset used and experimental procedure is provided.

The implementation of the experimental procedure was based on the Python
programming language, using the Scikit-Learn \cite{Pedregosa2011}, Imbalanced-
Learn \cite{JMLR:v18:16-365} and Geometric-SMOTE \cite{Douzas2019} libraries.
The experiments reported in this paper as well the analysis of the results are
reproducible using the scripts available at \url{https://github.com/AlgoWit/publications}.

\subsection{Baseline methods}

Four oversampling techniques were used in this experiment as baseline methods
along with Geometric SMOTE. Random oversampling was chosen for its simplicity.
SMOTE was selected for being a classic technique and the most popular
oversampling algorithm \cite{Douzas2019}. ADASYN \cite{HaiboHe2008} and
Borderline SMOTE \cite{Han2005} were selected for representing popular
modifications of the original SMOTE technique in the selection phase and data
generation mechanism, respectively. Finally, we use no oversampling as an
additional baseline method.

\subsection{Classification algorithms}

The selection of classification algorithms was done according to 3 criteria:
learning type, training time and popularity within the remote sensing
community. We further divide the relevant algorithms into 4 different learning
types: neighbours-based, rule-based, ensemble and generalised linear methods.

\todo{maybe move this part to the experimental setup?}
For each combination of oversampler and classifier (represented as $x_a$ and
$y_b$, respectively), a grid search will be run in order to find the optimal
parameter settings. The number of fits for a single combination  is defined as:

\[
fits_{x_a,y_b}=\prod\limits_{p=1} \textrm{count}(x_a^p).\prod\limits_{q=1}
\textrm{count}(y_b^q)
\]

Where $\textrm{count}(x_a^p)$ corresponds to the search grid size of
hyperparameter $p$ of an oversampler $x_a$. Finally, the total number of fits
is defined as:

\[
Total\ Fits=k\sum\limits_{a=1} \sum\limits_{b=1} fits_{x_a,y_b}
\]

Where $k$ is the number of folds defined for cross-validation. Thus, adding new
classifiers, oversamplers or grid search values may exponentially increase the
time necessary to run the experiment. So, in order to setup an extensive grid
search and run the full experiment within feasible time, selecting
computationally efficient classifiers becomes vital.

\cite{Khatami2016} developed a meta-analysis of classification processes out
of 266 manuscripts published from 1998 to 2012. Additionally,
\cite{Maxwell2018} also describe mature classification methods used in remote
sensing. The most common classifiers in supervised pixel-based land-cover image
classification processes include K-nearest neighbours, Support Vector Machines,
Decision Trees, Boosted Decision Trees, Random Forests and Neural Networks.
None of the manuscripts mention the use of generalised linear models.

Amongst the neighbours-based learning methods the K-nearest neighbours (KNN)
algorithm was selected. Out of the rule-based learning methods, the decision
tree classifier (DT) was chosen. Two ensemble learning methods were chosen:
Gradient boosting classifier and Random Forest classifier. Finally, the
logistic regression classifier was selected amongst the generalised linear
models, since it is a commonly used baseline algorithm. All these algorithms
were found to be computationally efficient and commonly used classifiers for
the proposed task (with exception for the logistic regression).

Despite their common use, both artificial neural networks and Support
Vector Machines were not included in this experiment due to their long training
times.

\subsection{Evaluation methods}

Amongst the many evaluation metrics existing for classifier's performance
evaluation, there are several indices that are considered to be preliminary and
are widely used for LULC classification accuracy assessment. Those measures are
classification overall accuracy, user's accuracy or precision and
producer's accuracy or recall,  and are calculated using the classification
confusion matrix \cite{Liu2007}. However, the LUCAS dataset has high
inter-class variability. Hence, the minority classes can be difficult to
account for by the classifiers that are designed to maximise the overall
accuracy, as the majority classes are receiving more weight \cite{Inglada2017}.
On the other hand, \cite{He2008} states that the use of user's and producer's
accuracy for class imbalanced data is not straightforward and producer's
accuracy is not sensitive to the data distribution at all. Considering these,
class specific metrics are needed to better describe classifier's performance. 
For this purpose, F-score and G-mean metrics are used to evaluate and compare 
the performance of the oversampling methods. Classification overall accuracy 
(OA) is provided for discussion.

\subsubsection{Overall Accuracy}

Classification Overall Accuracy is the number of correctly classified samples
divided by the sum of all the samples on the confusion matrix. 

In case when the training data is imbalanced and the accuracy metric is 
utilized, the result of the grid search for the best parameters will indicate 
that the particular parameterization is the one with highest classification 
overall accuracy, which is biased towards the majority class and can disregard 
the classification of minority classes.

\subsubsection{F-score}

F-score (or F-measure) is the harmonic mean of user's accuracy (precision) and
producer's accuracy (recall), where user's accuracy is the fraction of
correctly classified instances with regard to all instances classified as this
certain class in the confusion matrix, and producer's accuracy is the fraction
of  correctly classified instances with regard to all instances of that
reference class. The F-score value for multi-class imbalanced data
classification can be calculated considering the user's accuracy and producer's
accuracy as a unweighted average of the respective values from all the classes.

\[
F{-}score=2\frac{\text{E(UA)} \times \text{E(PA)}}{\text{E(UA)} + \text{E(PA)}}
\]

Where E(UA) and E(PA) are unweighted mean values of user's and producer's
accuracy of all the classes, respectively.

\subsubsection{G-measure}

The geometric mean score (G-mean) is the root of the product of class-wise
sensitivity and specificity. For the multi-class case sensitivity and
specificity are calculated for each class, then averaged with simple mean
value. Here the sensitivity (true positive rate) is the same with producer's 
accuracy (recall), and specificity (true negative rate) is defines as
the proportion of actual negatives classified as such. Adequately, specificity 
can be defined as producer's accuracy of the negative class \cite{Silva2017}.

\[
G{-}mean = \sqrt{E(Sensitivity) \times E(Specificity)}
\]

Where E(Sensitivity) and E(Specificity) are the unweighted mean values of
sensitivity and specificity of all the classes, respectively. The geometric 
mean indicates the balance between classification performances on both majority 
and minority classes. It means, high misclassification in the minority class 
will lead to a lower geometric mean value. This is a very valuable feature for 
evaluation of the classifier's performance with class imbalance problem, which 
is the case of this study. 


\subsection{Experiment Settings}

As it is stated in the section 3.2, machine-learning classifiers usually have
parameters that need to be set up by users for a better classification
performance. Parameter tuning to define the optimal values of the
hyper-parameters was performed with 5-fold cross-validation (CV) for each of
the model (combining classifiers with oversampler). The best parameters are
later selected to validate the classifier's performance.

In order to avoid from over-optimistic classification results, implementation
of CV with oversampling technique should be carried out as following
\cite{Lusa2015}. In $k$-fold CV dataset is divided into $k$ parts, one part is
excluded and used as only test set, while the rest $k$-1 part is used to train
the model. This process is iterated and each of the $k$ folds are used only
once as a test set to evaluate the classifier's performance. Later, results are
averaged across all $k$ iterations. It is crucial to remember that
all the steps for building the prediction model should be applied only on the
training data. From this respect, the oversampling (or any kind of data
resampling) technique should not be applied on the whole dataset. Instead, it
is used only on the training dataset generated in each step of the CV
procedure.

A range of hyper-parameters used for the grid search for each of the
classifier and oversampler are presented in the table \ref{tab:grid}.

\begin{table}[H]
	\centering
	\begin{tabular}{lll}
		\toprule
		Classifier       & Hyper-Parameters & Values\\
		\hline
		LR               & maximum iterations   & 10000   \\
		KNN              & number of neighbors  & {3, 5} \\
		DT               & maximum depth        & {3, 6} \\
		GBC              & maximum depth        & {3, 6} \\
    			 		 & number of estimators & {50, 100} \\
		RF               & maximum depth        & {None, 3, 6} \\
				 		 & number of estimators & {50, 100} \\
		\toprule
		Oversampler      &                      & \\
		\hline
		G-SMOTE          & number of neighbors  & {3, 5} \\
				 		 & selection strategy   & combined, minority, majority\\
				 		 & truncation factor    & {-1.0, -0.5, .0, 0.25,
				 		 0.5, 0.75, 1.0} \\
				 		 & deformation factor   & {0, 0.2, 0.4, 0.5,
				 		 0.6, 0.8, 1.0} \\
 		SMOTE            & number of neighbors & {3, 5} \\
		BORDERLINE SMOTE & number of neighbors & {3, 5} \\
		ADASYN           & number of neighbors & {2, 3} \\
		\bottomrule
	\end{tabular}
	\caption{\label{tab:grid}User-defined grid values}
\end{table}


\subsection{Data Set}

For this research LUCAS dataset and Landsat-8 time series from 2015 were used
for training and validation of supervised classifiers. 1694 ground-truth points 
from LUCAS 2015 dataset are used as reference data. This dataset is grouped 
into 8 classes and represents the main land cover types for the study area. The 
area of study is in the continental part of north-western Portugal, 
corresponding to the area covered by Landsat image from the track 204 and row 
32, with coordinates of center lat: 40.336, long: -8.468. \todo{not sure about 
the map of the study area, as there's not going to be a final map}

The dataset's class distribution and imbalance ratio is presented in Table 
\ref{tab:dataset_classes}. The data imbalance becomes visible with a minority 
class of 4 instances (wetlands) and majority class of 761 instances (woodland).

\begin{table}[H]
	\centering
	\begin{tabular}{llrrrr}
		\toprule
		\textbf{LUCAS Category} & \textbf{Land cover type} & \textbf{Instances}
		& \textbf{Imbalance Ratio} \\
		\hline
		A & Artificial land & 131 & 5.81 \\
		B & Cropland        & 270 & 2.81 \\
		C & Woodland        & 761 & 1.00 \\
		D & Shrubland       & 296 & 2.61 \\
		E & Grassland       & 185 & 4.11 \\
		F & Bareland        & 37  & 20.56 \\
		G & Water           & 10  & 76.10 \\
		H & Wetlands        & 4   & 190.25\\
		\hline
		\textbf{Total} & & \textbf{1694} &  \textbf{190.25} \\
		\bottomrule
	\end{tabular}
	\caption{\label{tab:dataset_classes}LUCAS nomenclature and sample distribution}
\end{table}

The remote sensing dataset presents 8 images from Landsat 8ETM+ multi-spectral
sensor. The images are Level-1 terrain corrected products (L1TP) and are
acquired in 2015 from February to September, 1 image each month. The 
acquisition mode is Descending. Only bands 2, 3, 4, 5, 6 and 7 are used from 
each image. Thus, each reference point from LUCAS dataset has 48 features 
(Table \ref{tab:datasets}). Those features have the value of digital numbers of 
the pixels from each spectral bands of the image time series representing LUCAS 
point location.

\begin{table}[H]
	\centering
	\csvautobooktabular{./resources/dataset_descrip.csv}
	\caption{\label{tab:datasets}Description of the dataset}
\end{table}

\section{Results and Discussion}
This section presents the final results of classifiers performance with 
different oversamplers over LUCAS dataset. A simple ranking of oversamplers 
performance is carried out for comparative analysis.  

The cross-validation scores for each combination of classifiers, evaluation 
metrics and oversamplers are presented in Table \ref{tab:mean_cross_val}. 
From this table we can observe that Geometric-SMOTE over-performs nearly all 
the other oversampling methods for both G-mean and F-score metrics with all the 
classifiers. 

\begin{table}[H]
	\centering
	\csvautobooktabular{./resources/mean_scores.csv}
	\caption{\label{tab:mean_cross_val}Results for cross validation scores 
		of oversamplers (B-SMOTE corresponds to Borderline SMOTE)}
\end{table}

A ranking score was assigned to each oversampling method with the best and 
worst performing methods receiving scores equal to 1 and 6, respectively. This 
comparison is presented in Table \ref{tab:ranking}. From this table we can 
observe that for G-mean measure Geometric-SMOTE algorithm systematically 
performs better than any other oversampling algorithm. For F-score it is the 
best for all the combinations but Borterline-SMOTE with Decision Tree 
classifier, even though, the difference with Geometric-SMOTE value is 
negligible.

\begin{table}[H]
	\centering
	\csvautobooktabular{./resources/mean_ranking.csv}
	\caption{\label{tab:ranking} Results for ranking of oversamplers}
\end{table}

The men of ranking scores for each of oversampling method across the 
classification algorithms is presented in the Table \ref{tab:mean_rankings}.

\begin{table}[H]
	\centering
	\csvautobooktabular{./resources/model_mean_ranking.csv}
	\caption{\label{tab:mean_rankings} Results for mean ranking of oversamplers}
\end{table}

As it was stated in the section 3.3, accuracy metric can be misleading for 
imbalanced data classification. This resulted the classification with 
unmodified data to have the highest accuracy score due to the fact, that 
overall accuracy is determined mostly by the majority classes, while importance 
of the prediction in minority classes is neglected. In the multi-class 
classification with imbalanced data such as LUCAS dataset, when the 
prediction of all the classes are of importance, accuracy metrics should not be 
implemented to evaluate the model performance, as is commonly utilized in 
remote sensing. However, even for accuracy metric, Geometric-SMOTE shows the 
best performance amongst the data modification algorithms and exceeds the rest 
of the oversampling methods.
 
Table \ref{tab:smote_gsmote} presents the percentage difference  between 
Geometric-SMOTE and SMOTE. For this comparison SMOTE algorithm was selected due 
to its popularity and wide use within the remote sensing society (see section 
2).

\begin{table}[H]
	\centering
	\csvautobooktabular{./resources/mean_perc_diff_scores.csv}
	\caption{\label{tab:smote_gsmote}Results for percentage difference between 
		G-SMOTE and SMOTE}
\end{table}

\section{Conclusions}



\bibliography{references}
\bibliographystyle{apalike}

\end{document}
